---
title: "Step 4 - Application to the Linearized Double-Logistic Function"
description: "All previous steps have been building to the application of Gibbs sampling for the double-logistic function."
author:
  - name: Matthew Shisler
    affiliation: North Carloina State University - Department of Statistics
    affiliation-url: https://statistics.sciences.ncsu.edu/ 
categories: [Bayesian, MCMC, Hierarchical Model] # self-defined categories
draft: false 
format:
  html: 
    code-fold: true
execute: 
  cache: true
  freeze: auto
---


```{r}
#| label: load-packages
#| output: false
#| code-summary: "Code: Load the packages"

library(tictoc)
library(Rfast)
library(numDeriv)
```

Let $Y_{ij}$ be an observation collected on day $t_{ij}$ of year $i$, $i = 1, \dots, N$ and $j = 1,\dots,n_i$. Assume $Y_{ij}$ to be normally distributed with mean $v(t_{ij}, \, \boldsymbol\theta_i)$ and constant variance $\sigma^2$. The mean "double-logistic" function, $v(t, \, \boldsymbol{\theta})$, is parameterized by $\boldsymbol\theta$ and is defined
\begin{equation}
\tag{1}
v(t, \, \boldsymbol\theta) = \theta_1 + (\theta_2 - \theta_7 t)\left(\frac{1}{\exp\left\{\frac{\theta_3 - t}{\theta_4}\right\}} - \frac{1}{\exp\left\{\frac{\theta_4 - t}{\theta_6}\right\}}\right)
\end{equation}

There is an additional constraint that $v(t, \, \boldsymbol{\theta}) \in [0,1]$ and $\theta_7 \ge 0$. One option would be to put priors on $\theta_1$ with $[0,1]$ support and $\theta_7$ with $[0,\infty)$ support. One might also consider the constraint $\theta_2 \in [\theta_1, 1]$, but we should be careful to specify a prior whose support depends on another parameter. Instead, we opt not to constrain or transform $\theta_2$. In general, we would prefer to place a multivariate normal prior over the entire vector $\boldsymbol\theta$. To that end, consider a reparameterazation of $v(t,\boldsymbol\theta)$,

\begin{equation}
\tag{2}
v(t, \, \boldsymbol\theta) = \frac{1}{1+\exp\{\theta_1\}} + \left(\theta_2 - \exp\{\theta_7\} t\right)\left(\frac{1}{\exp\left\{\frac{\theta_3 - t}{\theta_4}\right\}} - \frac{1}{\exp\left\{\frac{\theta_4 - t}{\theta_6}\right\}}\right)
\end{equation}

The function $v(t, \, \boldsymbol{\theta})$ is a non-linear function of $\boldsymbol\theta$ which consequently violates conjugacy of the typically assumed priors. To overcome this we will work with a linearized version of $v(t, \, \boldsymbol{\theta})$. This significantly decreases the computational burden of MCMC at the cost of introducing statistical bias. Are the computational gains worth the induced bias?

The linearization of $v(t, \, \boldsymbol\theta)$ around $\boldsymbol\theta_0$ yields,
\begin{equation}
\tag{3}
v(t, \, \boldsymbol\theta) \approx v(t, \, \boldsymbol\theta_0) + \nabla_{\boldsymbol\theta}v(t, \, \boldsymbol\theta)|_{\boldsymbol\theta=\boldsymbol\theta_0}(\boldsymbol\theta - \boldsymbol\theta_0) 
\end{equation}

Define the following,

\begin{equation}
\tag{4}
Y^*_{ij} = Y_{ij} - v(t_{ij}, \, \boldsymbol\theta_0)
\end{equation}

\begin{equation}
\tag{5}
X(t) = \nabla_{\boldsymbol\theta}v(t, \, \boldsymbol\theta)|_{\boldsymbol\theta=\boldsymbol\theta_0}
\end{equation}

\begin{equation}
\tag{6}
\boldsymbol\delta_i = \boldsymbol\theta_i-\boldsymbol\theta_0
\end{equation}


\begin{equation}
\mathbf{Y}_i \sim \text{Normal}_{n_i}\left(v(\mathbf{t}_i, \boldsymbol{\theta}_i), \; \sigma^2 \mathbf{I}_{n_i}\right)
\end{equation}

Construct the $m \times p$ parent design matrix $\mathbf{X}$ with rows $X(t)$ for $t=1,\dots,m$. Define the "child" design matrix, $\widetilde{\mathbf{X}}_i$ to be the subset of rows from $\mathbf{X}$ specified by $\mathbf{t}_i$. 

The $N \times q$ covariate matrix $\mathbf{Z}$ organizes known covariates for the mean of the distribution for $\boldsymbol\delta_i$. Let $\mathbf{z}_i$ be the $i$th row of $\mathbf{Z}$ and $\boldsymbol{\mathcal{Z}}_i = (\mathbf{z}^T_i\otimes \mathbf{I}_p)$ where "$\otimes$" represents the Kronecker product.

Now we can place a multivariate prior on $\boldsymbol\delta_i$ as this parameter represents deviations from $\boldsymbol\theta_0$. The linearized model is as follows,

\begin{align*}
\mathbf{Y}^*_i &\sim \text{Normal}_{n_i}\left(\widetilde{\mathbf{X}}_i\boldsymbol\delta_i, \; \sigma^2 \mathbf{I}_{n_i}\right)\\
\boldsymbol\delta_i &\sim \text{Normal}_p\left(\boldsymbol{\mathcal{Z}}_i\boldsymbol\beta, \; \mathbf{\Omega}\right)
\end{align*}

Again, we will assume $\mathbf{\Omega}$ is diagonal and let $\omega_{kk}$ be the $k$th diagonal element. Next we specify priors,
\begin{align*}
\boldsymbol\beta &\sim \text{Normal}_{qp}\left(\boldsymbol\mu, \; \mathbf{\Lambda}\right)\\
\omega_{kk} &\sim \text{InvGamma}\left(a_\omega, \; b_\omega \right)\\
\sigma^2 &\sim \text{InvGamma}\left(a_\sigma, \; b_\sigma\right)
\end{align*}

Let $\boldsymbol\Delta = [\boldsymbol\delta_1 \; \dots \; \boldsymbol\delta_N]$ and $\mathbf{B} = [\boldsymbol\beta_1 \; \dots \; \boldsymbol\beta_q]$. That is, $\mathbf{B}$ is a $p \times q$ matrix and $\text{vec}(\mathbf{B}) = \boldsymbol\beta$.  The full conditionals in this model are as follows,

\begin{align*}
\boldsymbol\delta_i|\text{rest} &\sim \text{Normal}_p(\mathbf{V}_i^{-1}\mathbf{M}_i, \mathbf{V}_i^{-1})\\
\mathbf{V}_i &= \frac{1}{\sigma^2} \widetilde{\mathbf{X}}_i^T\widetilde{\mathbf{X}}_i + \mathbf{\Omega}^{-1}\\
\mathbf{M}_i &= \frac{1}{\sigma^2} \widetilde{\mathbf{X}}_i^T\mathbf{Y}_i + \mathbf{\Omega}^{-1}\mathbf{B}\mathbf{z}_i\\\\
\boldsymbol\beta|\text{rest} &\sim \text{Normal}_{qp}(\mathbf{V}_\beta^{-1}\mathbf{M}_\beta, \mathbf{V}_\beta^{-1}\\
\mathbf{V}_\beta &= \mathbf{Z}^T\mathbf{Z} \otimes \boldsymbol\Omega^{-1} + \mathbf{\Lambda}^{-1}\\
\mathbf{M}_\beta &= (\mathbf{Z}^T \otimes \, \boldsymbol\Omega^{-1})\text{vec}(\boldsymbol\Delta) + \mathbf{\Lambda}^{-1}\boldsymbol\mu\\\\
\omega_{kk}|\text{rest} &\sim \text{InvGamma}(A_\omega,B_\omega)\\
A_\omega &= N/2 + a_\omega\\
B_\omega &= \frac{1}{2}\sum_{i=1}^N (\delta_{ik} - (\mathbf{B}\mathbf{z}_i)_k)^2 + b_\omega\\\\
\sigma^2|\text{rest} &\sim \text{InvGamma}(A_\sigma,B_\sigma)\\
A_\sigma &= \frac{1}{2}\sum_{i=1}^N n_i + a_\sigma\\
B_\sigma &= \frac{1}{2}\sum_{i=1}^N (\mathbf{Y}_i - \widetilde{\mathbf{X}}_i\boldsymbol\delta_i)^T(\mathbf{Y}_i - \widetilde{\mathbf{X}}_i\boldsymbol\delta_i) + b_\sigma
\end{align*}

Next moving on the code implementation. Define the double logistic function and its gradient.
```{r}
#| code-summary: "User-defined functions."

# expit #
# equivalent to plogis with m=0 & s=1.
expit   <- function(x){1/(1+exp(-x))}

# expit_p #
# first derivative of the expit function. equivalent to dlogis with m=0 & s=1.
expit_p <- function(x){expit(x)*(1-expit(x))}


double_logis <- function(t, theta){
  # double logistic function.
  # theta1 is transformed using the logistic function.
  # theta2 is transformed using the 
  # This allows for all parameters to follow a gaussian distribution
  
  theta[1] <- plogis(theta[1])
  theta[7] <- exp(theta[7])  
  
  n1 <- 1
  d1 <- 1 + exp((theta[3] - t)/theta[4])
    
  n2 <- 1
  d2 <- 1 + exp((theta[5] - t)/theta[6])
    
  out <- theta[1] + (theta[2] - theta[7]*t)*(n1/d1 - n2/d2)
  
  return(out)
}


# double logistic gradient wrt theta
basis_functions <- function(t,theta){
  t   <- t%%366

  dl0 <- double_logis(t,theta)-expit(theta[1])
  a <- expit_p(theta[7])
  theta[7] <- exp(theta[7])

  B1 <- expit_p(theta[1])
  B2 <- (1/(1+exp((theta[3]-t)/theta[4])))-(1/(1+exp((theta[5]-t)/theta[6])))
  B3 <- (theta[7]*t - theta[2])/(2*theta[4]*cosh((theta[3]-t)/theta[4])+2*theta[4])
  B4 <- ((theta[3] - t)*(theta[2]-theta[7]*t)*cosh((theta[3]-t)/(2*theta[4]))^(-2))/(4*theta[4]^2)
  B5 <- (theta[2] - theta[7]*t)/(2*theta[6]*cosh((theta[5]-t)/theta[6])+2*theta[6])
  B6 <- ((theta[5] - t)*(theta[7]*t-theta[2])*cosh((theta[5]-t)/(2*theta[6]))^(-2))/(4*theta[6]^2)
  B7 <- -t*a*((1/(1+exp((theta[3]-t)/theta[4])))-(1/(1+exp((theta[5]-t)/theta[6]))))
  B <- unname(cbind(B1, B2, B3, B4, B5, B6, B7))

  return(B)
}

```

We need to specify a value of $\boldsymbol\theta_0$ on which to center the linearization. We will expand on how to
obtain this value at a later time.
```{r}
theta0 <- c(-1.80, 0.75, 120, 8, 270, 8, -7.5)

plot(1:366, double_logis(1:366, theta0), type = "l",
     xlab = "t",
     ylab = bquote(v(t,theta[0])))
```

We verify that the analytical gradient is correct by checking the result against a numerical gradient function.
```{r}
#| code-summary: "Construct the numerical and analytical gradiant matrix"

theta0 <- c(-1.80, 0.75, 120, 8, 270, 8, -7.5)

gradinput <- function(x,t){
  return(double_logis(t,x))
}

# compute numerical gradient
G <- matrix(0, nrow = 366, ncol = 7)
for (i in 1:366){
  G[i,] <- grad(gradinput, x = theta0, t=i)
}

# # compute the analytical gradient
X <- basis_functions(1:366, theta0)

matplot(G, type="l", main = "Numerical Gradient", xlab = "t")
matplot(X, type="l", main = "Analytical Gradient", xlab = "t")

```

Next we simulate some data from the linearized model. (Need to add details regarding simulation set-up).
```{r}
#| code-summary: "Simulate the data"

# dimensions
N <- 40
m <- 366
n <- rep(100,N)
p <- 7
q <- 2

theta0 <- c(-1.80, 0.75, 120, 8, 270, 8, -7.5)

# Design matrices
gradinput <- function(x,t){
  return(double_logis(t,x))
}

# compute numerical gradient
Xp <- basis_functions(1:366, theta0)
Z  <- matrix(c(rep(1,N), seq(1:N)), ncol = 2)

# beta parameters
B0     <- matrix(c(0, 0, -15, 0, 15, 0, 0,
                   0, 0,   1, 0, -1, 0, 0), nrow=p)
beta0  <- matrix(c(B0), ncol = 1)
Omega0 <- diag(c(0.025, 1e-3,  3, 1,  3, 1, 0.0005))

# delta parameters
delta0  <- matrix(0, nrow = p, ncol = N)
sigma20 <- 0.0025

# sample data
Y <- list()
X <- list()
t <- list()
for (i in 1:N){
  # draw delta
  delta0[,i]   <- t(Rfast::rmvnorm(1, B0%*%Z[i,], Omega0))
  
  # draw rows from parent X
  t[[i]] <- sample(1:m, n[i])
  X[[i]] <- Xp[t[[i]],]
  
  # draw response
  Y[[i]] <- matrix(rnorm(n[i], mean = X[[i]]%*%delta0[,i], sd = sqrt(sigma20)), ncol=1)
}

```

Plot the simulated curves. We won't plot the individual observations.
```{r}
plot(1:366, double_logis(1:366, theta0), type = "l", ylim=c(-0.1, 1.2))

for (i in 1:N){
  lines(1:366, double_logis(1:366, theta0) + Xp%*%delta0[,i], type="l", col = i,
        xlab = "t",
        ylab = "Y",
        main = "Simulated Curves")
  # points(t[[i]], Y[[i]] + double_logis(t[[i]], theta0), pch=16)
}
```

The Gibbs loop. Later we will re-write this as a function.
```{r}
#| code-summary: "Run the Gibbs sampler"

# set-up
niter <- 5000
keep_delta  <- array(NA, dim = c(p, N, niter))
keep_B      <- array(NA, dim = c(p, q, niter))
keep_Omega  <- matrix(NA, nrow = niter, ncol = p)
keep_sigma2 <- rep(NA, niter)

# initial values
delta  <- matrix(0, nrow = p, ncol = N)
beta   <- rep(0, q*p)
B      <- matrix(c(beta), nrow=p)
sigma2 <- 0.01
Omega  <- diag(c(0.5, 0.5, 5, 0.5, 5, 0.5, 0.001))
keep_delta[,,1] <- delta
keep_B[,,1]     <- B
keep_Omega[1,]  <- diag(Omega)
keep_sigma2[1]  <- sigma2

# prior parameters
mu    <- rep(0, q*p)
Lambda_inv <- diag(rep(1e-06,q*p))
a     <- 0.1
b     <- 0.1

# pre-computes
XtX <- list()
XtY <- list()
for (k in 1:N){
  XtX[[k]] <- t(X[[k]])%*%X[[k]]
  XtY[[k]] <- t(X[[k]])%*%Y[[k]]
}
ZtZ <- t(Z)%*%Z
Lmu <- Lambda_inv%*%mu
Ao    <- N/2 + a
As    <- sum(n)/2 + a

# Gibbs Loop
tic()
for (iter in 2:niter){

  Omega_inv <- diag(1/diag(Omega))
  
  # sample deltas
  for (i in 1:N){
    M         <- (1/sigma2)*XtY[[i]] + Omega_inv%*%(B%*%Z[i,])
    V_inv     <- chol2inv(chol((1/sigma2)*XtX[[i]] + Omega_inv))
    delta[,i] <- V_inv%*%M+t(chol(V_inv))%*%rnorm(p)
  }
  
  # sample beta
  M     <- kronecker(t(Z), Omega_inv)%*%matrix(c(delta), ncol = 1) + Lmu
  V_inv <- solve(kronecker(ZtZ, Omega_inv) + Lambda_inv)
  beta  <- V_inv%*%M+t(chol(V_inv))%*%rnorm(q*p)
  B     <- matrix(beta, nrow = p)
  
  # sample omegas
  for (k in 1:p){
    Bo <- sum((delta[k,] - (B%*%t(Z))[k,])^2)/2 + b
    Omega[k,k] <- 1/rgamma(1, Ao, Bo)
  }
  
  # sample sigma2
  SSE <- 0
  for (i in 1:N){
    SSE <- SSE + sum((Y[[i]] - X[[i]]%*%delta[,i])^2)
  }
  Bs <- SSE/2 + b
  sigma2 <- 1/rgamma(1, As, Bs)
  
  # store everything
  keep_delta[,,iter] <- delta
  keep_B[,,iter]     <- B
  keep_Omega[iter,]  <- diag(Omega)
  keep_sigma2[iter]  <- sigma2
}
toc()


```

Now for some trace plots. The following are for $\mathbf{B}$, $\boldsymbol\Omega$, and $\sigma^2$. And display iterations 100:5000. This visual inspection seems to indicate good convergence!
```{r}
#| code-summary: "Construct trace plots"
win <- 1:niter

par(mfrow = c(2,2))

for(l in 1:q){
  for (k in 1:p){
    plot(win, keep_B[k, l, win], type = "l",
         ylab = bquote(beta[paste(.(l),",",.(k))]),
         xlab = "iter")
    abline(h = B0[k, l], col = "red")
  }
}

for (k in 1:p){
  plot(win, keep_Omega[win, k], type = "l",
       ylab = bquote(omega[paste(.(k),",",.(k))]),
       xlab = "iter")
  abline(h = Omega0[k, k], col = "red")
}

plot(win, keep_sigma2[win], type = "l",
     ylab = bquote(sigma^2),
     xlab = "iter")
abline(h = sigma20, col = "red")

```

Trace plots for 5 subject $\boldsymbol\delta_i$ chosen at random.
```{r}
#| code-summary: "Construct trace plots"

win <- 1:niter

par(mfrow = c(2,2))
param_sample <- sample(1:N, 5)

for (year in param_sample){
  for (k in 1:p){
    subscr <- paste0(year,",",k)
    plot(win, keep_delta[k, year, win], type = "l",
         ylab = bquote(delta[.(subscr)]),
         xlab = "iter")
    abline(h = delta0[k, year], col = "red")
  }
}

```
