---
title: "ST746 - Computer Assignment 1"
description: ""
author:
  - name: Matthew Shisler
    affiliation: North Carloina State University - Department of Statistics
    affiliation-url: https://statistics.sciences.ncsu.edu/ 
categories: [Stochastic Proccesses, ST 746] # self-defined categories
draft: false 
format:
  html: 
    code-fold: true
---



## Step 1

Create a vector of transition probabilities from a single state.
```{r}
num.states <- 8

t1 <- runif(num.states)
t1 <- t1/sum(t1)

round(t1,2)
```

## Step 2

Create the entire transition matrix.
```{r}
P <- matrix(runif(num.states^2), nrow = num.states)
P <- P/rowSums(P)

```

## Step 3

Simulate the Markov chain with initial state $X_0 = 1$. Store the proportion of time the chain was in state $j$, $j=1,\dots,8$.
```{r}


```

## Step 4

Simulate the Markov chain with initial states $X_0 = 2,\dots,8$. For each initial state, store the proportion of time the chain was in state $j$, $j=1,\dots,8$.

```{r}


```


## Step 5

Compare the proportion vectors from questions 3 and 4.

```{r}


```

## Step 6

Compute the $P^{100}$ matrix multiplication. Are its rows approx equal? How do they compare the row vectors from step 5?

```{r}


```


## Step 7

Solve the equation of stationarity $\boldsymbol\mu = \boldsymbol\mu \mathbf{P}$. Is the solution nearly equal to that in step 6?

```{r}


```

## Step 8


Write your own conclusions based on this simulation experiment and give an explanation of the phenomenon you observe.
```{r}


```








