---
title: "ST746 - Computer Assignment 1"
description: ""
author:
  - name: Matthew Shisler
    affiliation: North Carloina State University - Department of Statistics
    affiliation-url: https://statistics.sciences.ncsu.edu/ 
categories: [Stochastic Proccesses, ST 746] # self-defined categories
draft: false 
format:
  html: 
    code-fold: false
---

```{r, include=FALSE}
library(expm)
library(MASS)
```

## Step 1

Create a vector of transition probabilities from a single state.
```{r}
num.states <- 8

t1 <- runif(num.states)
t1 <- t1/sum(t1)

round(t1,2)
```

## Step 2

Create the entire transition matrix. Note this actually throws out the first vector we
created in step 1, but fulfills the overall spirit of the exercise.
```{r}
P <- matrix(runif(num.states^2), nrow = num.states)
P <- P/rowSums(P)
P
```

## Step 3

Simulate the Markov chain with initial state $X_0 = 1$. Store the proportion of time the chain was in state $j$, $j=1,\dots,8$.
```{r}
current.state <- 1
num.steps <- 10000
visits1 <- rep(0, num.states)

for (i in 1:num.steps){

  visits1[current.state] <- visits1[current.state] + 1
  current.state <- sample(1:num.states, 1, prob = P[current.state,])
    
}

round(visits1/sum(visits1),3)

```

## Step 4

Simulate the Markov chain with initial states $X_0 = 2,\dots,8$. For each initial state, store the proportion of time the chain was in state $j$, $j=1,\dots,8$.

```{r}
visits2    <- matrix(0, nrow=num.states-1, ncol=num.states)
num.steps <- 10000

for (j in 1:(num.states-1)){
  current.state <- j
  
  for (i in 1:num.steps){

    visits2[j, current.state] <- visits2[j, current.state] + 1
    current.state <- sample(1:num.states, 1, prob = P[current.state,])
    
  }
  
}

round(visits2/rowSums(visits2),3)

```


## Step 5

Compare the proportion vectors from questions 3 and 4.

```{r}

boxplot(visits2/rowSums(visits2))
points(visits1/sum(visits1), col = "red", cex = 8, pch = ".")
```

## Step 6

Compute the $P^{100}$ matrix multiplication. Are its rows approx equal? How do they compare the row vectors from step 5?

```{r}
P %^% 100

```


## Step 7

Solve the equation of stationarity $\boldsymbol\mu = \boldsymbol\mu \mathbf{P}$. Is the solution nearly equal to that in step 6?

```{r, warning=F}

e    <- eigen(P)
lvec <- MASS::ginv(e$vectors)

as.numeric(lvec[1,]/sum(lvec[1,]))

```

## Step 8


Write your own conclusions based on this simulation experiment and give an explanation of the phenomenon you observe.










