---
title: "Basics of Timeseries AR1 Models"
description: "Starting from the basics. . ."
author:
  - name: Matthew Shisler
    affiliation: North Carloina State University - Department of Statistics
    affiliation-url: https://statistics.sciences.ncsu.edu/ 
categories: [Research, Bayesian, MCMC] # self-defined categories
draft: false 
format:
  html: 
    code-fold: true
---


```{r}
#| label: load-packages
#| output: false
#| code-summary: "Code: Load the packages"
```


This will be a short description of the basic auto-regressive lag-1 model in time series analysis.

Consider a simple situation where a researcher makes $n$ measurements $y_1,\dots,y_n$, on some scientific object of interest and further suppose that we wish to model these observations as independent and identically distributed Gaussian random variables $Y_i$ with unknown mean $\mu$ and variance $\sigma^2$,

$$
Y_i \sim \text{N}(\mu, \sigma^2)
$$

Estimating $\mu$ and $\sigma^$ is straightforward enough by maximizing the likelihood and applying the usual bias correction:

$$
\widehat\mu = \frac{1}{n}\sum_{i=1}^n y_i \quad \quad \quad \widehat\sigma^2 = \frac{1}{n-1}\sum_{i=1}^n (y_i - \widehat\mu)^2
$$

In reality these measurements were most likely collected sequentially and indexed by time $t$, $Y_t$, $t = 1, \dots, n$. Depending on the nature of the scientific object being measured it is not unreasonable to suspect that a measurement made at time $t$ may be correlated with a measurement made at time $t-1$ and it would be wise for the scientific researcher to account for this fact in their statistical analysis.

To illustrate the issue let's consider an example where we simulate some iid data and some correlated data.

```{r}
n  <- 1000
mu <- 0
sd <- 1

y.iid <- rnorm(n, mu, sd)

plot(y.iid)
plot(density(y.iid))
```

```{r}
y.ts <- arima.sim(list(ar=0.99),n=n)

plot(y.ts)
plot(density(y.ts))

```







